Design Document for R Package glmm

### 2016-06-02 NewDesignDoc/DesignDocGLMM.tex
Minor changes to design doc and pkg's manual
Expanding ntrials arg to likelihood level
Previously only cumulant function had arg ntrials. Now elc, elval, elGH all take ntrials arg. Tests and calls to these functions updated to reflect this.

\title{Design Document: Monte Carlo Maximum Likelihood for Generalized Linear Mixed Models}

@@

For the Bernoulli distribution, we calculate these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) as follows. In order to be careful with our computer arithmetic, we use the \texttt{log1p} function and the second set of equalities.  
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}
 For Poisson, these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

@@

\subsection{Equations}
Recall the log of the data density is
\begin{align}
\log f_\theta(y|u) &= y' \eta +c(\eta) \\
&= \sum_{i} y_{i} {\eta_{i}} - c({\eta_{i}})
\end{align}
where
\begin{align}
\eta=X\beta+Zu.
\end{align}
The derivative of this with respect to one component, $\eta_j$, is
\begin{align}
\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = y_j-c'(\eta_j).
\end{align}
The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l}$, is
\begin{align}
\dfrac{\partial \eta_j}{\partial \beta_{l}} = X_{j{l}}
\end{align}

@@

@@

&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }- l(\theta)  \right] \\

@@

\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }- \sqrt{m}\;  l(\theta)  \\

@@

 - l(\theta) \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   

>>>>

\title{Design Document for R Package glmm}

@@

For the Bernoulli distribution, we calculate these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) as follows. In order to be careful with our computer arithmetic, we use the \texttt{log1p} function and the second set of equalities.  
\begin{align}
c(\eta_i) &= \log(1+e^{\eta_i}) =
  \begin{cases}
    \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
    \eta_i+\log(e^{-\eta_i}+1) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{1}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{1}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}

 For Poisson, these values (\texttt{cum}, \texttt{cp}, and \texttt{cpp}) are
\begin{align}
c(\eta_i)&=e^{\eta_i}\\
c'(\eta_i)&=e^{{\eta_{i}}}\\
c''(\eta_i)&= e^{{\eta_{i}}}.
\end{align}

@@

\subsection{Equations}
Recall the log of the data density is
\begin{align}
\log f_\theta(y|u) &= y' \eta -c(\eta) \\
&= \sum_{i} y_{i} {\eta_{i}} - c({\eta_{i}})
\end{align}
where
\begin{align}
\eta=X\beta+Zu.
\end{align}
The derivative of this with respect to one component, $\eta_j$, is
\begin{align}
\dfrac{\partial}{\partial \eta_j} \log f_\theta(y|u)  = y_j-c'(\eta_j).
\end{align}
The derivative of the component $\eta_j$ with respect to one of the fixed effect predictors, $\beta_{l}$, is
\begin{align}
\dfrac{\partial \eta_j}{\partial \beta_{l}} = X_{j{l}}
\end{align}

@@

\section{New Additions (Summer 2016)}
\subsection{New Family: Binomial}
One year later, I'm adding binomial. For Binomial, the calculations change only in the cumulant function. They rely on one more number: the number of trials. The inputted response is usually a vector, but for binomial (with the exception of Bernoulli) we need 2 vectors cbinded (cbound?) together. The first column is the number of successes and the second is the number of failures. Summing across the row gives us the number of trials. I call this $ntrials$ from here on out.

The cumulant function and its first two derivatives with respect to the canonical parameter $\eta_i$ are below. 
\begin{align}
c(\eta_i) &=ntrials_i \, \log(1+e^{\eta_i}) =
  \begin{cases}
   ntrials_i \, \log(1+e^{\eta_i}) & \text{if } \eta_i\leq 0,\\
   ntrials_i \,\left(  \eta_i+\log(e^{-\eta_i}+1) \right) & \text{if } \eta_i >0,
  \end{cases}\\
c'(\eta_i)&=\dfrac{ e^{{\eta_{i}}}}{ 1+e^{{\eta_{i}}}} = \dfrac{ntrials_i}{1+e^{-\eta_i}}\\
c''(\eta_i)&=   \dfrac{e^{{\eta_{i}}}}{  1+ e^{{\eta_{i}}} }  - \dfrac{e^{2{\eta_{i}}}}{    (  1+ e^{{\eta_{i}}})^2}  = \dfrac{ntrials_i}{1+e^{-\eta_i}}\cdot \dfrac{1}{1+e^{\eta_i}}
\end{align}

I will implement this change in steps. Game plan:
\begin{enumerate}
\item Start by writing the binomial cumulant in R. Check to make sure this calculation is identical to that produced by bernoulli for some parameter value $\eta_i$ and for $ntrials=1$. Check for both $\eta$ positive and negative.
\item Check the three calculations for $ntrials=5$ (or some number other than 1) compared to my by-hand calculation. 
\item Code it in C and check that the calculation is identical to that produced in R for $ntrials=1$ and for $ntrials \neq 1$.
\item The second lowest level will then call binomial.glmm rather than bernoulli.glmm as the family. I will work my way up the levels incorporating ntrials. At the top, I will have a variable to set ntrials = 1 (for now). As I do each layer, I will make sure I get all the same answers from my salamanders problem (or some other bernoulli problem).
\item I will go to the top-most layer, right where the model matrices are being made. The new response vector y will be the first column of the user-provided response. The sum of the two entries in the row will create the vector ntrials. I will check that the salamanders problem gives the same results (when specified as cbind(Mate, 1-Mate)).

\item I will add checks for the model matrix stage.  See below.
\item I want to allow the response for bernoulli to be a single vector  (rather than two cbinded vectors). Therefore, if the family is binomial.glmm and the length of the dimension of the response is 1 (meaning we just have a vector), then create the ntrials vector to be a vector of 1s. Test this with salamanders or some other problem to make sure it gives the same results.

\end{enumerate}

Checks for the model matrix stage: 
\begin{enumerate}
\item  If  the length of the dimension of the response is 2 dimensions (meaning we have both rows and columns), then the family must be binomial.glmm.
\item If the family is binomial.glmm, then the response must have 2 columns
\end{enumerate}

These changes will need to be done in both C and R. The checks will be in R. I'll need to add  documentation for binomial. Bernoulli will be retained.

@@

&\sqrt{m} \left[  \dfrac{\dfrac{1}{m}\sum_{k=1}^m    \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)}  }- \nabla l(\theta)  \right] \\

@@

\dfrac{\dfrac{1}{\sqrt{m}}\sum_{k=1}^m     \nabla \log f_\theta(u_k,y)   \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }- \sqrt{m}\; \nabla  l(\theta)  \\

@@

 -\nabla l(\theta) \dfrac{1}{\sqrt{m}}\sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }{\dfrac{1}{m} \sum_{k=1}^m  \dfrac{f_\theta(u_k,y)}{\tilde{f}(u_k)} }   
